---
title: "LTR-RT death rates"
author: "cstritt"
date: "18/07/2019"
output: html_document
---

# Two methods to estimate the death rates of LTR retrotransposon families

#### From Material & Methods:
Assuming that full-length copies die at a constant rate, the survival func-
tion of a LTR-RT family can be approximated by an exponential function
of the form Nt = N0e−λt , where λ is the death rate of the family. To esti-
mate λ for the nine families with at least 20 full length copies, exponential
distributions were fitted to the age distributions using the fitdistr function
of the R package MASS, and the exponential rate was recovered. In ad-
dition, we estimated λS , an alternative death rate estimate which includes
information on the age of solo LTRs, using the maximum likelihood func-
tion of Dai et al. (2018). Confidence intervals for λS were obtained from 100
bootstraps.

```{r}
setwd(dir = "/home/valentin-grenet/Bureau/Données/TE_sequences/")
```

```{r, include=TRUE, eval=FALSE}

library(stats4)
library(mosaic)

# Implement the MLE approach of Dai et al. 2018, p. 1043
deathrate <- function(y, age, lambda){
  full = y*exp(-lambda*age)
  solo = (1-y)*(1-exp(-lambda*age))
  return(solo + full)
}
l_joint <- function(lambda){
  R <- mapply(deathrate, df$y, df$estimated_age, lambda)
  -log(prod(R))
}


# Load table
tg <- read.table('/home/valentin-grenet/Bureau/Données/Phoenix_dactylifera.full_length_all_ages.tsv', header=T, dec=",")

# Only use high-confidence annotated TEs
# tg <- subset(tg, (cat == 'single' & tsd == 1 & ltr_context == 'solo_LTR') | 
#                   (cat == 'paired' & tsd == 1 & strand == '+' & ltr_context == "five_prime_LTR-three_prime_LTR") |
#                   (cat == 'paired' & tsd == 1 & strand == '-' & ltr_context == "three_prime_LTR-five_prime_LTR"))


# How many observations per family?
table(tg$consensus)
age_df <- data.frame()
lambda_estimates <- data.frame()

# Select families with at least ...
setwd(dir = "/home/valentin-grenet/Bureau/Données/TE_sequences/")
pwd <- getwd()
consensi <- list.files(path = "./", pattern = "consensus")
include <- c('consensus_Cluster_256_subfam_2', 'consensus_Cluster_226', 'consensus_Cluster_113', 'consensus_Cluster_53_subfam_2')

for (te in include){ 
  leaf_ages <- subset(tg, tg$consensus == te)
  if (nrow(leaf_ages)<10) next

  # encode solo LTRs as 0 and FLEs as 1
  leaf_ages$y <- rep(NA, nrow(leaf_ages))
  
  leaf_ages$y[grepl("single", leaf_ages$LTR)] <- 0
  leaf_ages$y[grepl("solo", leaf_ages$LTR)] <- 0
  leaf_ages$y[grepl("complete", leaf_ages$LTR)] <- 1
  leaf_ages$y[grepl("paired", leaf_ages$LTR)] <- 1
  leaf_ages$y[grepl("shrunked", leaf_ages$LTR)] <- 1
  leaf_ages$y[grepl("fragmented", leaf_ages$LTR)] <- 1
    
  # remove rows with missing data
  rmv <- which(is.na(leaf_ages$y))
  if (length(rmv) > 0){
    leaf_ages <- leaf_ages[-rmv, ]
  }
    
  
  # First approach: fit geometric distribution to age distribution and recover death rate
  leaf_ages_fle <- subset(leaf_ages, y==1)
  geom1 <- MASS::fitdistr(leaf_ages_fle$estimated_age, "geometric")
  geom1$loglik <- sum(stats::dgeom(leaf_ages_fle$estimated_age, geom1[["estimate"]], 
                                     log = TRUE))
    
  p1 <- geom1$estimate["prob"]
  deathRate <- p1/(1 - p1) # eq.(4)?
    
  # # Second approach, including solo LTR age
  # df <- leaf_ages
  # fit <- mle(l_joint, start=list(lambda=1), method="Brent", lower = 0.001, upper = 20)
  # leaf_ages$survival <- exp(-fit@coef*leaf_ages$estimated_age)
  #   
  # # bootstrapping
  # nboot = 100
  # bootstrap <- c()
  # for (i in 1:nboot){
  #   df <- leaf_ages[sample(nrow(leaf_ages), nrow(leaf_ages), replace=T), ]
  #   fit <- mle(l_joint, start=list(lambda=1), method = "Brent", lower = 0.001, upper = 20)
  #   bootstrap <- c(bootstrap, unname(fit@coef))
  # }
  # qtls<- qdata(bootstrap, c(0.025, 0.975))
  
  # extend data frames
  age_df <- rbind(age_df, leaf_ages)
    
  # lambda_row <- data.frame('family' = te, 'lambda'=unname(deathRate),'lambdaS' = unname(fit@coef),
                           # 'lower' = qtls[1], 'upper' = qtls[2])
  lambda_row <- data.frame('family' = te, 'lambda'=unname(deathRate))
    
  lambda_estimates <- rbind(lambda_estimates, lambda_row)
}

lambda_estimates$half_life <- log(2) / lambda_estimates$lambda
lambda_estimates$half_life_S <- log(2) / lambda_estimates$lambdaS
lambda_estimates$half_life_CI_l_S <- log(2) / lambda_estimates$lower
lambda_estimates$half_life_CI_u_S <- log(2) / lambda_estimates$upper

write.table(lambda_estimates, file = '../Phoenix_dactylifera.complete_lambdas.tsv', row.names = FALSE, col.names = TRUE, append = TRUE, sep = "\t", dec = ".")

ggplot(age_df) + 
  geom_smooth(aes(x=estimated_age, y=survival, col=consensus)) 


# Plot survival functions
survival <- data.frame()
time = seq(0,10, 0.01)

for (i in 1:nrow(lambda_estimates)){
  
  x <- lambda_estimates[i,]
  prob <- exp(-x$lambda*time)
  # prob_ci_l <- exp(-x$lower*time)
  # prob_ci_u <- exp(-x$upper*time)

  te <- rep(x$family, length(prob))
  acc <- rep(x$acc, length(prob))
  
  df <- data.frame(time, te, prob)#, prob_ci_l, prob_ci_u)
  survival <- rbind(survival, df)

}

ggplot(survival,aes(x=time, y=prob, col=te))  + 
  geom_line(linewidth=1) +
  theme_bw(base_size=12, base_family = "Arial") +
  xlab("Time (million years)") + ylab("Survival probability")+
  scale_color_viridis_d(option='D')
```
